---
title: "forest_fall_data_practice_exploring"
format: html
---

This is for me to practice playing with data from Katy Trail Fall 2023. For the sake of my sanity at the moment
I will just be dealing with the 12pm data. 3 pumps were used, pump #3, pump #4, and pump #5. There was also one control tube that was not opened during the sampling. I have also included the three blanks that were at the beginning of the run to give a baseline of machine noise.

This leaves us with 

```{r}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
```


```{r}
#reading the data from M365_help_write_function.qmd

blk_1_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/blk_1_ff23")
blk_2_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/blk_2_ff23")
blk_3_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/blk_3_ff23")
cntrl_12pm_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/control_12pm_ff23")
pmp_3_12pm_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/p3_12pm_ff23")
pmp_4_12pm_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/p4_12pm_ff23")
pmp_5_12pm_ff23 <- read.csv("../practice_tdu_data/combined_csv_ff23/p5_12pm_ff23")


```

```{r}
#attempting to make new columns to better categorize and track the data
#added the columns "location", "pump", "time", and "year". The blks and controls
#don't have pumps so they are NA, and same for blk time and location.
pmp_3_12pm_ff23_org <- pmp_3_12pm_ff23 |> 
  mutate(location = "kt", pump_num = 3, time = "12pm", year = "2023")

pmp_4_12pm_ff23_org <- pmp_4_12pm_ff23 |> 
  mutate(location = "kt", pump_num = 4, time = "12pm", year = "2023")

pmp_5_12pm_ff23_org <- pmp_5_12pm_ff23 |> 
  mutate(location = "kt", pump_num = 5, time = "12pm", year = "2023")

cntrl_12pm_ff23_org <- cntrl_12pm_ff23 |> 
  mutate(location = "kt", pump_num = NA, time = "12pm", year = "2023")

blk_1_ff23_org <- blk_1_ff23 |> 
  mutate(location = NA, pump_num = NA, time = NA, year = "2023")

blk_2_ff23_org <- blk_2_ff23 |> 
  mutate(location = NA, pump_num = NA, time = NA, year = "2023")

blk_3_ff23_org <- blk_3_ff23 |> 
  mutate(location = NA, pump_num = NA, time = NA, year = "2023")

```

Combining these into one df
```{r}
ff_12pm_23 <- bind_rows(pmp_3_12pm_ff23_org, pmp_4_12pm_ff23_org, pmp_5_12pm_ff23_org)
```


Reading in the pump count info

```{r}
pump_count_ff23 <- read_csv("Tidy_Pump_Counts_ff23.csv") # read in file with pump counts
```

Adding calculations to this pump count

```{r}
pump_count_ff23 <- pump_count_ff23 |>
  mutate(total_count = count_after-count_before) #count after minus count before to get total counts during sampling

pump_count_ff23 <- pump_count_ff23 |>
  mutate(total_ml = total_count*ml_count) #total ml collected

pump_count_ff23 <- pump_count_ff23 |>
  mutate(total_l = total_ml/1000) #total l collected

pump_count_ff23 <- pump_count_ff23 |>
  mutate(cubic_m = total_ml/1000000) #total cubic meters collected
```

Joining the pump info to the compound info
```{r}
ff_12_23_joined <- left_join(ff_12pm_23, pump_count_ff23, by=join_by(time, pump_num))
```

Corrected concentrations based on the pump volume (ng from gc/meters cubed)
```{r}
ff_12_23_corrected <- ff_12_23_joined |>
  mutate(ng_cubic_m = Amount/cubic_m) #adding a new column with ng/cubic meter concentration
```

```{r}
#let's try some basic visualization!

ff_12pm_23 |> 
  group_by(Peak_Name) |>
  summarize(avg_conc = mean(Amount), std_conc = sd(Amount)) |>
  print(n=35)

```

```{r}
ff_12_23_corrected |>
  ggplot(aes(x= Peak_Name, y= ng_cubic_m, color = Peak_Name)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 1) +
  theme( axis.text.x = element_text(
    angle = 90
  ))
```

```{r}
#trying to log transform the data to spread it apart and show the relationships better
ff_12_23_corrected <- ff_12_23_corrected |>
  mutate(ng_cubic_m_log = log(ng_cubic_m))

ff_12_23_corrected |>
  ggplot(aes(x= Peak_Name, y= ng_cubic_m_log, color=Peak_Name)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 1) +
  theme( axis.text.x = element_text(
    angle = 90
  ))
```

```{r}
ff_12_23_corrected |>
  summarize(avg_conc = mean(ng_cubic_m)) #views the mean of the average phytoncide amount

c_range <- ff_12_23_corrected |>
  group_by(Peak_Name) |>
  summarize(average_compound = mean(ng_cubic_m))

in_range <- c_range |>
  subset(average_compound<100) |>
  subset(average_compound>0.5)
#After averaging the ng_cubic_m for each compound then filtering against the top and bottom of the calibration curve, only 27 of them fall within the range. The others are extrapolation.

out_range <- c_range |>
  subset(average_compound>100 | average_compound<0.5)
#these are the compounds beyond the range of calibration

```

```{r}
#filtering just alpha pinene and p cymene

ff_12_23_filtered <- ff_12_23_corrected  |>
  filter(Peak_Name %in% c("p Cymene", "Alpha Pinene"))

print(ff_12_23_filtered)

ff_12_23_filtered |>
  ggplot(aes(x= Peak_Name, y= ng_cubic_m, color = Peak_Name)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 1)
```



Now we want to create a table for publication that has 
- compound name
- RT
- Primary ion
- Secondary ion
- Tertiary ion
- Line equation
- R^2
- LOD
- LOQ

I think I need to filter the data down a bit then export it so I can merge it to the table in the LOD/LOQ

```{r}
ion_info <- ff_12_23_corrected |>
  select(Peak_Name, Expected_Retention_Time, Target_Ion, Qualifier1_Ion, Qualifier2_Ion )

ion_info <- unique(ion_info) #dropping the duplicate columns

ion_info <- ion_info |>
  rename(name = Peak_Name)
```

