---
title: "many_xmls_to_tibble"
format: html
---

This was written with the help of Microsoft CoPilot using the following prompts
"Using the code below I can combine 4 xmls into one tibble, however I would like to combine many xmls into one big tibble. The xml names and their files help identify where they came from and this is important for keeping track of where the data is from. 

Modify this code so I can put in file named "data" and there are 6 files with names like "phytoncide_collection_como_fall_2023" and within each of those 6 files there are 16 files that have names such as "p5_1800_forest_fall_2023", "blk_4_forest_fall_2023", "control_1800_forest_fall_2023", etc. and within each of the 16 files are the 4 xml files labeled like "p3_1800_mix1", "p3_1800_mix2", etc. The code should be able to take these xml files and convert them into a tibble and add identifcation keys based on the file names they come from such as "forest", "spring", "2023", "p3", "control", "blk". The code should also be transferrable so it can handle having different numbers of files added if we ever change the sampling technique."

insert libraries
--------------------------------------------
```{r}
library(xml2)
library(dplyr)
library(purrr)
library(tibble)
library(readr)
library(stringr)
```


--------------------------------------------
Function to turn xml into tibble

```{r}
xml_compounds_to_tibble <- function(path,
                                   record_xpath = ".//Compounds",
                                   col_types = NULL,
                                   add_source = TRUE) {
  doc <- read_xml(path)

  nodes <- xml_find_all(doc, record_xpath)

  rows <- map(nodes, function(node) {
    kids <- xml_children(node)
    nms  <- xml_name(kids)
    vals <- xml_text(kids)

    # Normalize empty strings to NA
    vals <- na_if(str_trim(vals), "")

    as.list(setNames(vals, nms))
  })

  out <- bind_rows(rows) |> as_tibble()

  # Convert types if provided (keeps unspecified columns as character)
  if (!is.null(col_types)) {
    out <- type_convert(out, col_types = col_types, na = c("", "NA"))
  }

  if (add_source) {
    out <- mutate(out, source_file = basename(path), .before = 1)
  }

  out
}
```

--------------------------------------------
Extracts Meta-Data from Paths and Filenames

```{r}
extract_metadata <- function(path) {

  path_parts <- str_split(path, .Platform$file.sep)[[1]]

  # folder levels
  top_level  <- path_parts[length(path_parts) - 3]  # e.g. phytoncide_collection_como_fall_2023
  location   <- path_parts[length(path_parts) - 2]  # e.g. p5_1800_forest_fall_2023
  xml_file   <- tools::file_path_sans_ext(basename(path))

  tibble(
    collection = top_level,

    # tokens from intermediate folder
    pump     = str_extract(location, "p\\d+"),
    treatment  = case_when(
      str_detect(location, "control") ~ "control",
      str_detect(location, "blk")     ~ "blk",
      TRUE ~ NA_character_
    ),
    environment = str_extract(location, "forest|field|mix"),
    season      = str_extract(location, "spring|summer|fall|winter"),
    year        = str_extract(location, "\\d{4}"),

    # tokens from xml file
    mix         = str_extract(xml_file, "mix\\d+"),
    replicate   = str_extract(xml_file, "rep\\d+|mix\\d+")
  )
}
```


```{r}
extract_metadata2 <- function(path) {

  path_parts <- str_split(path, .Platform$file.sep)[[1]]

  project_dir <- path_parts[length(path_parts) - 3]
  sample_dir  <- path_parts[length(path_parts) - 2]

  xml_name <- tools::file_path_sans_ext(basename(path))

  # infer sample type
  sample_type <- case_when(
    str_detect(sample_dir, "^p\\d+")    ~ "sample",
    str_detect(sample_dir, "^blk")      ~ "blk",
    str_detect(sample_dir, "^control") ~ "control",
    TRUE ~ NA_character_
  )

  tibble(
    project = project_dir,

    sample_type = sample_type,

    pump = if_else(
      sample_type == "sample",
      str_extract(sample_dir, "p\\d+"),
      NA_character_
    ),

    time = str_extract(sample_dir, "(?<=_)\\d{3,4}(?=_)|(?<=_)\\d{3,4}$"),

    location = str_extract(sample_dir, "como|forest|field"),

    season = str_extract(sample_dir, "spring|summer|fall|winter"),

    year = str_extract(sample_dir, "\\d{4}"),

    mix = str_extract(xml_name, "mix\\d+"),
    
    




```

Testing what the heck this is doing so I can parce it correctly
```{r}

extract_metadata3 <- function(path) {
  
path_parts <- str_split(path, .Platform$file.sep)[[1]]

file_dir <- path_parts[length(path_parts) - 3]
project_dir  <- path_parts[length(path_parts) - 2]
sample_dir <- path_parts[length(path_parts)- 1]
xml_name <- tools::file_path_sans_ext(basename(path))

  # infer sample type
  sample_type <- case_when(
    str_detect(sample_dir, "^p\\d+")    ~ "sample",
    str_detect(sample_dir, "^blk")      ~ "blk",
    str_detect(sample_dir, "^control") ~ "control",
    TRUE ~ NA_character_
  )

  tibble(
    project = project_dir,

    type = sample_type,

    pump = if_else(
      sample_type == "sample",
      str_extract(sample_dir, "p\\d+"),
      NA_character_),

    time = if_else(
      sample_type == "sample",
      str_extract(sample_dir, "(?<=_)\\d{3,4}(?=_)|(?<=_)\\d{3,4}$"),
      NA_character_),

    location = str_extract(project_dir, "como|forest|field"),

    season = str_extract(project_dir, "spring|summer|fall|winter"),

    year = str_extract(project_dir, "\\d{4}"),

    mix = str_extract(xml_name, "mix\\d+"),
    
    sample_ID <- case_when(
      
    sample_type == "sample" ~ str_c(
      "sample",
      pump,
      time,
      location,
      season,
      year,
      mix,
      sep = "_",
      na.rm = TRUE
    ),

    # Controls (time + year usually meaningful)
    sample_type == "control" ~ str_c(
      "control",
      time,
      mix,
      location,
      season,
      year,
      sep = "_",
      na.rm = TRUE
    ),

    # Blanks (often only numeric batch + mix)
    sample_type == "blk" ~ str_c(
      "blk",
      time,
      mix,
      location,
      season,
      year,
      sep = "_",
      na.rm = TRUE
    ),

    TRUE ~ NA_character_
  )

    )
}
```

```{r}
extract_metadata4 <- function(path) {

  path_parts <- str_split(path, .Platform$file.sep)[[1]]

  file_dir    <- path_parts[length(path_parts) - 3]
  project_dir <- path_parts[length(path_parts) - 2]
  sample_dir  <- path_parts[length(path_parts) - 1]

  xml_name <- tools::file_path_sans_ext(basename(path))

  # ---- sample type ----
  sample_type <- case_when(
    str_detect(sample_dir, "^p\\d+")    ~ "sample",
    str_detect(sample_dir, "^blk")      ~ "blk",
    str_detect(sample_dir, "^control") ~ "control",
    TRUE ~ NA_character_
  )

  # ---- shared metadata ----
  pump <- if_else(
    sample_type == "sample",
    str_extract(sample_dir, "p\\d+"),
    NA_character_
  )

   time = if_else(
      sample_type %in% c("sample", "control"),
      str_extract(sample_dir, "(?<=_)\\d{3,4}(?=_)|(?<=_)\\d{3,4}$"),
      NA_character_
      )
   
  location <- str_extract(project_dir, "como|forest|field")
  season   <- str_extract(project_dir, "spring|summer|fall|winter")
  year     <- str_extract(project_dir, "\\d{4}")
  mix      <- str_extract(xml_name, "mix\\d+")
  blk_num <- str_extract(sample_dir, "blk_\\d+")

  # ---- sample ID (type-aware) ----
  sample_ID <- case_when(

    sample_type == "sample" ~ str_c(
      "sample", pump, time, location, season, year, mix,
      sep = "_"
    ),

    sample_type == "control" ~ str_c(
      "control", time, location, season, year, mix,
      sep = "_"
    ),

    sample_type == "blk" ~ str_c(
      "blk", location, season, year, mix,
      sep = "_"
    ),
    TRUE ~ NA_character_)

  tibble(
    project   = project_dir,
    type      = sample_type,
    pump      = pump,
    blk_num   = blk_num,
    time      = time,
    location  = location,
    season    = season,
    year      = year,
    mix       = mix,
    sample_ID = sample_ID
  )
}
```

--------------------------------------------
Reading all xml files

```{r}
read_all_xmls <- function(root_dir,
                          record_xpath = ".//Compounds",
                          col_types = NULL) {

  xml_files <- list.files(
    root_dir,
    pattern = "\\.xml$",
    recursive = TRUE,
    full.names = TRUE
  )

  message("Found ", length(xml_files), " XML files")

  map_dfr(xml_files, function(path) {

    compounds <- xml_compounds_to_tibble(
      path,
      record_xpath = record_xpath,
      col_types = col_types,
      add_source = TRUE
    )

    meta <- extract_metadata4(path)

    bind_cols(
      meta %>% slice(rep(1, nrow(compounds))),
      compounds
    )
  })
}
```

--------------------------------------------
column type assignment

```{r}
compound_types <- cols(
  Number                  = col_integer(),
  Peak_Name               = col_character(),
  Retention_Time          = col_double(),
  Amount                  = col_double(),
  Peak_Area               = col_double(),
  Units                   = col_character(),
  Cal_Istd                = col_integer(),
  QValue                  = col_integer(),
  Expected_Retention_Time = col_double(),
  Target_Ion              = col_double(),
  Qualifier1_Ion          = col_double(),
  Qualifier2_Ion          = col_double(),
  Qualifier3_Ion          = col_double(),
  Start_Integration       = col_double(),
  End_Integration         = col_double()
)


```

--------------------------------------------
put file here

```{r}
foresttest <- read_all_xmls(
  root_dir = "data test/",
  col_types = compound_types
)

glimpse(foresttest)
#write_csv(all_compounds, "all_phytoncide_compounds.csv")
```

