---
title: "lod_loq_calculation_practice"
format: html
---
Example:
Standard concentration: 10 ppm benzene
Peak height: 5 pA
Noise: 0.03 pA
Detection threshold: 3× noise

set up relationship
10ppm <- 5pA
LOD <- 3 X noise

General Formula:
LOD = (Factor × Noise × Concentration) / Peak Height

Applying the numbers:
LOD = (3 × 0.03 pA × 10 ppm) / 5 pA
LOD = 0.18 ppm

So any benzene peak above 0.18 ppm is considered detectable. Anything lower is unreliable and usually treated as “not detected.”

LOQ:
LOQ = 3.3 x LOD



___________________________
signal = peak height
noise = in the TIC what is the nearby background noise
concentration = in ppm the concentration looked at
injection_ul = Injection volume (usually 1uL)
line equation: response = slope * amount

```{r}
#loading libraries

library(tidyverse)
library(readxl)
library(gtable)
library(dplyr)
library(ggpmisc)
```

Reading in the Data
```{r}
#reading in table with all the cal curve info and signal to noise

cal_info <- read_xlsx("compound_summary_table_info_2024.xlsx")

glimpse(cal_info)

knitr::kable(cal_info)

```

Adding Columns for LOD and LOQ
```{r}
#LOD = (Factor × Noise × Concentration) / Peak Height

cal_info <- cal_info |>
  mutate(lod = (3*concentration)/(signal/noise),
         loq = (10*concentration)/(signal/noise))

glimpse(cal_info)
```

Adding Line Equation Row
```{r}
cal_info <- cal_info |>
  mutate(line = paste("y =", line_slope, "x"))

glimpse(cal_info)
```


Merging data frames and selecting/renaming columns
```{r}
ion_table_dat <- left_join(cal_info, ion_info)

ion_table_dat <- ion_table_dat |>
  select(name, Expected_Retention_Time, Target_Ion, Qualifier1_Ion, Qualifier2_Ion,line, R2_cal_curve, lod, loq) |>
  rename("Compound" = name, "RT" = Expected_Retention_Time, "Target Ion" = Target_Ion, "Q1 Ion" = Qualifier1_Ion, "Q2 Ion" = Qualifier2_Ion, "line" = line, "R^2" = R2_cal_curve, "LOD" = lod, "LOQ" = loq)

ion_table_dat <- ion_table_dat |>
  arrange(RT) #sorting by increasing RT 

glimpse(ion_table_dat)
```


Creating Table
```{r}
library(knitr)
library(kableExtra)

```

```{r}

ion_table <- kable(ion_table_dat, caption = "Ion Info", digits = 3) %>%
  kable_styling(full_width = FALSE, position = "center", 
                latex_options = c("striped", "hold_position"))

ion_table
```

Finding LOD min and max
```{r}
#ion_table_dat |>
 #summarise(LOD_min = min(ion_table_dat$LOD),
  #          LOD_max = max(ion_table_dat$LOD),
   #         LOQ_min = min(ion_table_dat$LOQ),
    #        LOQ_max = max(ion_table_dat$LOQ))
```

```{r}
#used AI to help write this based on above code
ion_table_dat |>
  summarise(
    LOD_min       = min(LOD, na.rm = TRUE),
    LOD_min_names = str_c(Compound[LOD == LOD_min], collapse = ", "),
    LOD_max       = max(LOD, na.rm = TRUE),
    LOD_max_names = str_c(Compound[LOD == LOD_max], collapse = ", "),
    LOQ_min       = min(LOQ, na.rm = TRUE),
    LOQ_min_names = str_c(Compound[LOQ == LOQ_min], collapse = ", "),
    LOQ_max       = max(LOQ, na.rm = TRUE),
    LOQ_max_names = str_c(Compound[LOQ == LOQ_max], collapse = ", ")
  )

```

```{r}
#LOD under 1pmm

lod_under1 <- filter(ion_table_dat, ion_table_dat$LOD<1)

lod_over1 <-filter(ion_table_dat, ion_table_dat$LOD>1)
```


```{r}

ion_table_dat |>
  summarise(mean_R2 = mean(ion_table_dat$`R^2`),
            min_R2 = min(`R^2`),
            max_R2 = max(`R^2`))


```

Is slope correlated with LOD?
```{r}
cal_info |>
  ggplot(aes(x= lod, y= line_slope)) +
  geom_point() +
  scale_x_reverse()

#No...it really doesn't
```

```{r}
#creating calibration curve plot of Camphene

camphene_conc <- c(0.5, 1, 10, 25, 50, 75, 100) #in ppm
camphene_resp <- c(550, 752, 23740, 62877, 117078, 196228, 248727) #peak area

camphene_cal_curv <- data.frame(
  concentration = camphene_conc,
  response = camphene_resp
)


camphene_cal_curv |>
ggplot(aes(x=concentration, y= response)
       ) +
  stat_poly_eq(formula = y~x+0,
               parse = TRUE,
               rr.digits = 4
               ) +
  stat_poly_line(formula = y~x+0,
                 se =FALSE,
                 color = "black"
                 )+
  geom_point(size = 4,
             alpha = .7
             ) +
  labs(title = "Camphene Calibration Curve",
       x = "Concentration (ppm)",
       y = "Peak Area")
  
```

```{r}

camphene_conc <- c(0.5, 1, 10, 25, 50, 75, 100) #in ppm
camphene_resp <- c(550, 752, 23740, 62877, 117078, 196228, 248727) #peak area

camphene_cal_curv <- data.frame(
  concentration = camphene_conc,
  response = camphene_resp
)


camphene_cal_curv |>
ggplot(aes(x=concentration, y= response)
       ) +
  stat_poly_eq(formula = y~x+0,
               parse = TRUE,
               rr.digits = 4
               ) +
  stat_poly_line(formula = y~x+0,
                 se =FALSE,
                 color = "black"
                 )+
  geom_point(size = 4,
             alpha = .7
             ) +
  labs(title = "Camphene Calibration Curve",
       x = "Concentration (ppm)",
       y = "Peak Area")
  
```

