---
title: "phytoncide_collection_working"
format: html
---

The goal for this document is to:
- get the file loaded in
- load in tidy pump data
- use that pump data to ge the corrected concentration readings
- once the corrected concentration info is in, check pump performance
    - do this through general visualization faceting and coloring
    - then start doing relationship plots to see if certain compounds and pumps
    the reason some of the readings are so high and variable.
    
# Load in ya libraries :)
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
theme_set(theme_cowplot())
library(factoextra)
library(wesanderson)
library(knitr)
library(gt)
library(openxlsx)
pal <- wes_palette(name = "Zissou1")
```

# Read in ya data!
```{r}
phyto_dat <- read.csv("phytoncide_dat")

phyto_dat$time <- factor(phyto_dat$time, labels = c("8am", "12pm", "6pm"), levels = c("800", "1200", "1800")) #changing time to factor 

glimpse(phyto_dat)

pump_dat <- read_xlsx("tidy_pump_counts.xlsx")

pump_dat$time <- factor(pump_dat$time, labels = c("8am", "12pm", "6pm"), levels = c("800", "1200", "1800"))

glimpse(pump_dat)
```

# Performing calculations on the pump data
```{r}
pump_dat <- pump_dat |>
  mutate(total_count = count_after-count_before, #pump count during sampling
         total_ml = total_count*ml_count, #ml collected
         total_l = total_ml/1000, #total liters collected
         cubic_m = total_ml/1000000) #total cubic meters collected

glimpse(pump_dat)
```

# Slap those files together
```{r}
dat_joined <- left_join(phyto_dat, pump_dat, by=join_by(time, pump, year, season, location)) 
#joined on the shared columns

glimpse(dat_joined)
```

# Calculate the corrected concentrations based on the pump data and added ID column for PCA analysis and ug/m^3
```{r}
dat <- dat_joined |>
  mutate(conc_ngcm = amount/cubic_m,
         conc_ugcm = conc_ngcm/1000,
         pca_id = case_when(

    type == "sample" ~ str_c(
      "sample", pump, time, location, season, year,
      sep = "_"
    ),

    type == "control" ~ str_c(
      "control", time, location, season, year,
      sep = "_"
    ),

    type == "blk" ~ str_c(
      blk_num, location, season, year,
      sep = "_"
    ),
    TRUE ~ NA_character_))
```
```{r}
datcheck <- dat |>
  filter_out(type == "control") |>
  filter_out(type == "blk") |>
  group_by(peak_name) |>
  summarise(n = n())
```

## Okay now we should be good to go, and we can start visualization now!

giant scatter plot of all the data points faceted and colored for different features
```{r}
dp <- dat |>
  ggplot(aes(x = season, y = conc_ngcm, color = pump)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5)

dp
```

histogram of data
```{r}
dphis <- dat |>
  ggplot(aes(x = conc_ngcm)) +
  geom_histogram(bins = 30)

dphis
```

histogram faceting
```{r}
dphisf1 <- dat |>
  ggplot(aes(x = conc_ngcm)) +
  geom_histogram(bins = 30) +
  facet_wrap(~season)

dphisf1
```

density plot of data
```{r}
dpden <- dat |>
  ggplot(aes(x = conc_ngcm)) +
  geom_density()

dpden
```

scatterplot with location and pump
```{r}
fc <- dat |>
  ggplot(aes(x = location, y = conc_ngcm, color = pump)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5)
fc
```

scatterplot looking at pumps
```{r}
fc2 <- dat |>
  ggplot(aes(x = pump, y = conc_ngcm, color = location)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5)
fc2
```

scatterplot looking at compounds
```{r}
fc3 <- dat |>
  ggplot(aes(x = peak_name, y = conc_ngcm, color = pump)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 1) +
  theme( axis.text.x = element_text(
    angle = 90))
fc3
```

scatterplot looking at time
```{r}
fc4 <- dat |>
  ggplot(aes(x = time, y = conc_ngcm, color = location)) +
  geom_point(position = position_jitter(width = 0.5),
             alpha = 0.5,
             size = 3) +
  theme( axis.text.x = element_text(
    angle = 90))
fc4
```

What do we see from this?

- The data is highly right skewed (plots dpden and dphis)
- data points from the city on beta-phellandrene are a major cause of this skew (plot fc3)
- It appears that the pumps have a similar spread of data, so the pumps are 
all functioning the equally (plot fc2)
- you can't really see any major differences between the season in this form (fc)


Where do we go from here?

- log transformation
- recheck skew and spread

# Log Transformation Time!!!

```{r}
dat <- dat |>
  mutate(conc_nglog = log10(conc_ngcm),
         conc_uglog = log10(conc_ugcm))
```

Checking the skew again...
```{r}
dphis2 <- dat |>
  ggplot(aes(x = conc_uglog)) +
  geom_histogram(bins = 30)

dphis2
```

```{r}
dpden2 <- dat |>
  ggplot(aes(x = conc_uglog, fill = location)) +
  geom_density()

dpden2
```

```{r}
dphisf2 <- dat |>
  ggplot(aes(x = conc_nglog)) +
  geom_histogram(bins = 30) +
  facet_wrap(~season + location)

dphisf2
```

Let's see those plots again but this time they are readable >:)

scatterplot with location and pump
```{r}
fcl <- dat |>
  ggplot(aes(x = location, y = conc_nglog, color = pump)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5)
fcl
```

scatterplot looking at pumps
```{r}
fc2l <- dat |>
  ggplot(aes(x = pump, y = conc_nglog, color = location)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5)
fc2l
```

scatterplot looking at compounds
```{r}
fc3l <- dat |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = pump)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 1) +
  theme( axis.text.x = element_text(
    angle = 90))
fc3l
```

scatterplot looking at time
```{r}
fc4l <- dat |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90))
fc4l
```

Filtering some data out to look at only
```{r}
gp <- dat |>
  filter(type!="blk" & type!="control") |> #filtering out control and blanks
  ggplot(aes(x = season, y = conc_nglog)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 1)
gp


```


# Average Pumps no transformation
```{r}
dat_mean <- dat |>
  filter(type!="blk" & type!="control") |>
  group_by(location, season, time, peak_name) |>
  summarise(mean = mean(conc_ngcm), sd = sd(conc_ngcm))
```

histogram of this is still skewed
```{r}
mhis <- dat_mean |>
  ggplot(aes(x = mean)) +
  geom_histogram(bins = 30)

mhis
```

# Log Transformation of Averaged Pump info
```{r}
dat_mean <- dat_mean |>
  mutate(mean_log = log10(mean))
```

```{r}
mhis2 <- dat_mean |>
  ggplot(aes(x = mean_log)) +
  geom_histogram(bins = 30)

mhis2
```

visualization of this
```{r}
mfc <- dat_mean |>
  ggplot(aes(x = location, y = mean_log, color = time)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90))
mfc
```

# Selecting then visualizing
```{r}
#forest spring selection
fsp <- dat |>
  filter(type == "sample" & location == "forest" & season == "spring" & mix == "mix1")

fsp |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 2) + 
  facet_wrap(~time) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "Forest Spring Mix 1")
```

Interaction Plot of Forest Spring
```{r}
fsp |>
  ggplot(aes(x = time, y = conc_nglog, group = peak_name, color = peak_name )) +
   geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.5)
```

# Selecting the compounds that were significant from metaboanalyst
```{r}

top5 <- dat |>
  filter(type == "sample" & peak_name %in% c("limonene", "ocimene", "alpha_pinene", "beta_phellandrene", "sabinene"))
```

```{r}
top5 |>  
  ggplot(aes(x = location, y = conc_nglog, color = peak_name)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90))
```

mean then log of top 5
```{r}
top5m <- dat_mean |>
  filter(peak_name %in% c("limonene", "ocimene", "alpha_pinene", "beta_phellandrene", "sabinene"))

top5m |>  
  ggplot(aes(x = location, y = mean_log, color = peak_name)) +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_grid(season ~ time) +
  theme( axis.text.x = element_text(
    angle = 90))
```

# At this point I'm drowning in figures so I asked AI to walk me through some steps to visualize things
Boxplot by factor
```{r}
dat |>
  ggplot(aes(x = location, y = conc_nglog, fill = season)) +
  geom_boxplot() +
  facet_wrap(~peak_name, scales = "free_y")
```

Time of Day Trends
```{r}
dat |>
  filter(type == "sample", location == "como", season == "fall") |>
  ggplot(aes(x = time, y = conc_nglog, group = season, color = season )) +
  geom_point(position = position_jitter(width = 0.1, height = 0), alpha = 0.5) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar") +
  stat_summary(fun = mean, geom = "line") +
  facet_wrap(~peak_name, scales = "free_y") +
  labs(title = "Como Fall")

```

```{r}
dat |>
  filter(type == "sample", location == "forest", season == "fall") |>
  ggplot(aes(x = time, y = conc_nglog, group = season, color = season )) +
  geom_point(position = position_jitter(width = 0.1, height = 0), alpha = 0.5) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar") +
  stat_summary(fun = mean, geom = "line") +
  facet_wrap(~peak_name, scales = "free_y") +
  labs(title = "Forest Fall")

  
```

# PCA analysis!!!

Rearranging the data for PCA
```{r}
wide_dat <- dat |>
  filter(type == "sample") |>
  select(peak_name, conc_ngcm, pca_id) |>
  drop_na() |>
  pivot_wider(names_from = peak_name, values_from = conc_ngcm) |>
  select(-pca_id)
```

Running the PCA
```{r}
z <- prcomp(~.,
            data = wide_dat,
            center = TRUE,
            scale. = TRUE)

```

getting eigenvalues
```{r}
get_eig(z)

```

Printing the PCA

```{r}
print(z)
```

Visualize PCA

PCA 1 and 2
```{r}
pca1 <- fviz_pca_var(z)

ggsave(filename = "pca1.png", plot = pca1, width = 8, height = 7, units = "in", bg = "white")
```

PCA 1 and 2
```{r}
fviz_pca_var(z, axes = c(1, 2))
```

getting the PCA loadings as a tibble
```{r}
pcatibble <- as_tibble(
  z$rotation,
  rownames = "compound"
)
```

Now filtering based on the PCA to make  some graphs that are a bit easier to look at.
```{r}
pca1top12 <- dat |>
  filter(peak_name %in% c("limonene", "beta_myrcene", "alpha_pinene", "beta_phellandrene", "sabinene", "beta_pinene", "eucalyptol", "alpha_phellandrene", "alpha_terpinene", "gamma_terpinene", "longifolene", "ocimene"))

pca1top12 |>
  filter(type == "sample", location == "como", season == "fall" ) |>
  ggplot(aes(x = time, y = conc_nglog, group = season, color = season )) +
  geom_point(position = position_jitter(width = 0.1, height = 0), alpha = 0.5) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar") +
  stat_summary(fun = mean, geom = "line") +
  facet_wrap(~peak_name, scales = "free_y") +
  labs(title = "Como Fall")


```

making two plots to look at these compounds over time and separating by location
```{r}
pca1forest <- pca1top12 |>
  filter(type == "sample", location == "forest", peak_name == "limonene") |>
  ggplot(aes(x = time, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "Forest")
pca1forest

pca1city <- pca1top12 |>
  filter(type == "sample", location == "como", peak_name == "limonene") |>
  ggplot(aes(x = time, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "CoMo")
pca1city

plot_grid(pca1forest, pca1city, nrow = 2)
```

```{r}
blimonene <- pca1top12 |>
  filter(type == "sample", peak_name == "limonene") |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "both limonene")

blimonene

```

```{r}
bbetaphell <- pca1top12 |>
  filter(type == "sample", peak_name == "beta_phellandrene") |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "both alpha_phellandrene")

bbetaphell
```

```{r}
bmyrc <- pca1top12 |>
  filter(type == "sample", peak_name == "beta_myrcene") |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "both beta myrcene")

bmyrc
```

```{r}
sab <- pca1top12 |>
  filter(type == "sample", peak_name == "sabinene") |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "both sabinene")

sab
```

```{r}
ocim <- pca1top12 |>
  filter(type == "sample", peak_name == "ocimene") |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "both ocimene")

ocim
```

```{r}
apin <- pca1top12 |>
  filter(type == "sample", peak_name == "alpha_pinene") |>
  ggplot(aes(x = time, y = conc_nglog, color = location)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.3),
             alpha = 0.5,
             size = 3) +
  facet_wrap(~season) +
  theme( axis.text.x = element_text(
    angle = 90)) +
  labs(title = "both alpha_pinene")

ocim
```

Interaction plot of 5 compounds (keep this idiot)
```{r}
intplotpca <- pca1top12 |>
  filter(type == "sample", location == "forest", peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = time, y = conc_nglog, group = peak_name, color = peak_name )) +
   geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  facet_wrap(~season) +
  labs(title = "Forest", x = NULL, y = "Concentration (ng/m^3 log10)", color = NULL)

intplotpca

ggsave(filename = "interactionplotforest.png", plot = intplotpca, width = 8, height = 7, units = "in", bg = "white")

```

```{r}
intplotpca2 <- pca1top12 |>
  filter(type == "sample", location == "como", peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = time, y = conc_nglog, group = peak_name, color = peak_name )) +
   geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  facet_wrap(~season) +
  labs(title = "City", x = NULL, y = "Concentration (ng/m^3 log10)", color = NULL)

intplotpca2

ggsave(filename = "interactionplotcity.png", plot = intplotpca2, width = 8, height = 7, units = "in", bg = "white")
```

```{r}
plot_grid(intplotpca, intplotpca2, nrow = 2)
```

Plotting the Compounds next to each other 

```{r}
springplot1 <- pca1top12 |>
  filter(type == "sample", season == "spring", time == "12pm" , peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(0.3), size = 2, alpha = 0.5) +
  facet_wrap(~location) +
  theme(axis.text.x = element_blank()) +
  labs(title = "12pm")

springplot1

springplot2 <- pca1top12 |>
  filter(type == "sample", season == "spring", time == "8am" , peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(0.3), size = 2, alpha = 0.5) +
  facet_wrap(~location) +
  theme(axis.text.x = element_blank()) +
  labs(title = "8am")

springplot2

springplot3 <- pca1top12 |>
  filter(type == "sample", season == "spring", time == "6pm" , peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(0.3), size = 2, alpha = 0.5) +
  facet_wrap(~location) +
  theme(axis.text.x = element_blank()) +
  labs(title = "6pm")

springplot3

plot_grid(springplot1, springplot2, springplot3, nrow = 2)
  
```

Plotting the 12pms next to each other
```{r}
noonplot1 <- pca1top12 |>
  filter(type == "sample", season == "spring", time == "12pm" , peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(0.3), size = 2, alpha = 0.5) +
  facet_wrap(~location) +
  theme(axis.text.x = element_blank()) +
  labs(title = "Spring 12pm", x = NULL, color = NULL, y = "Concentration (ng/m^3 log10)")

noonplot1

noonplot2 <- pca1top12 |>
  filter(type == "sample", season == "summer", time == "12pm" , peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(0.3), size = 2, alpha = 0.5) +
  facet_wrap(~location) +
  theme(axis.text.x = element_blank()) +
  labs(title = "Summer 12pm", x = NULL, color = NULL, y = "Concentration (ng/m^3 log10)")

noonplot2

noonplot3 <- pca1top12 |>
  filter(type == "sample", season == "fall", time == "12pm" , peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = peak_name, y = conc_nglog, color = peak_name)) +
  geom_boxplot() +
  geom_point(position = position_jitter(0.3), size = 2, alpha = 0.5) +
  facet_wrap(~location) +
  theme(axis.text.x = element_blank()) +
  labs(title = "Fall 12pm", x = NULL, color = NULL, y = "Concentration (ng/m^3 log10)")

noonplot3

plot_grid(noonplot1, noonplot2, noonplot3, nrow = 2)

```

```{r}
fvc <- pca1top12 |>
  filter(type == "sample", peak_name %in% c("alpha_pinene", "ocimene", "sabinene", "beta_phellandrene", "limonene")) |>
  ggplot(aes(x = location, y = conc_nglog)) +
  geom_point(position = position_jitter(0.2), size = 2, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "royalblue") +
  stat_summary(fun.data = mean_se, geom = "errorbar", color = "royalblue", width = 0.2) +
  scale_x_discrete(labels = c("City", "Forest")) +
  labs(y = "Concentration (ng/m^3 log10)", x = "Location" )


fvc


  
```

```{r}
# ggsave(filename = "forestandcityscat.png", plot = fvc, width = 8, height = 7, units = "in", bg = "white")
# 
# ggsave(filename = "noonspring.png", plot = noonplot1, width = 8, height = 7, units = "in", bg = "white")
# 
# ggsave(filename = "noonsummer.png", plot = noonplot2, width = 8, height = 7, units = "in", bg = "white")
# 
# ggsave(filename = "noonfall.png", plot = noonplot3, width = 8, height = 7, units = "in", bg = "white")


```


Exporting a table to put into metaboanalyst
```{r}
metabo <- dat |>
  filter(type == "sample") |> 
  group_by(pump, season, location, peak_name) |>
  summarise(daymean = mean(conc_ugcm)) |>
  pivot_wider(names_from = peak_name, values_from = daymean)


#write.csv(metabo, file = "metabo_dat_phyto2.csv")

metabo2 <- dat |>
  filter(type == "sample") |> 
  group_by(time, season, location, peak_name) |>
  summarise(daymean = mean(conc_ugcm)) |>
  pivot_wider(names_from = peak_name, values_from = daymean)

write.csv(metabo2, file = "metabo_dat_phyto_time.csv")

metabo_forest <- dat |>
  filter(type == "sample", location == "forest") |> 
  group_by(pump, season, peak_name) |>
  summarise(daymean = mean(conc_ugcm)) |>
  pivot_wider(names_from = peak_name, values_from = daymean)
  
write.csv(metabo_forest, file = "metabo_dat_phyto_forest.csv")

metabo_como <- dat |>
  filter(type == "sample", location == "como") |> 
  group_by(pump, season, peak_name) |>
  summarise(daymean = mean(conc_ugcm)) |>
  pivot_wider(names_from = peak_name, values_from = daymean)

write.csv(metabo_como, file = "metabo_dat_phyto_como.csv")


```

```{r}
#checking to make sure this is working properly.
#it is indeed, my manually calcuated average matches the one from the metabo table
limonene <- dat |> filter(pump == "p5", location == "forest", season == "fall", peak_name == "limonene")

limonene2 <- dat |> filter(time == "12pm", location == "forest", season == "fall", peak_name == "limonene")
```



Hi, I am getting some average compound info and slapping it in a table to share with kintek

```{r}
mean_ugcm <- dat |>
  filter(type == "sample") |>
  group_by(location, peak_name) |>
  summarise(mean_conc = mean(conc_ugcm), stdv = sd(conc_ugcm))

kable(mean_ugcm, caption = "Microgram per Cubic Meter Mean of Each Compound")

write.xlsx(as.data.frame(mean_ugcm), "micrograms_per_cubic_meter.xlsx")



#freakin wizard with R now hehe
```


Asking the question: Which compounds should we choose to move forward into exposure testing
Things to consider:
- abundance of the compounds
- prior lit on health outcomes
- ease of access (can I purchase them easily)
- compliments other literature


To answer which compounds are most prevalent in the forest let's plot that
```{r}
#filtered by samples and forest as location. Then grouped by compound to find the average concentration of the compounds in the forest.

fcomp <- dat |>
  filter(type == "sample", location == "forest") |>
  group_by(peak_name) |>
  summarise(mean = mean(conc_ugcm))

ccomp <- dat |>
  filter(type == "sample", location == "como") |>
  group_by(peak_name) |>
  summarise(mean = mean(conc_ugcm))
```

> the compounds with the highest concentration are 
